---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(tibble.print_min = 6, tibble.print_max = 6)
```

# weibulltools

<!-- badges: start -->
[![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![CRAN status](https://www.r-pkg.org/badges/version/weibulltools)](https://CRAN.R-project.org/package=weibulltools)
<!-- badges: end -->

## Overview

The *weibulltools* package focuses on statistical methods and visualizations that are often used in reliability engineering.  

The goal of *weibulltools* is to equip the user with a compact and easily accessible set of methods and visualization tools that make the examination and adjustment as well as the analysis and interpretation of field data (and bench tests) as simple as possible.  

Besides the well-known weibull analysis, *weibulltools* supports multiple lifetime distributions and also contains Monte Carlo methods for the correction and completion of imprecisely recorded or unknown lifetime characteristics.  
Plots are created statically ([ggplot2](https://ggplot2.tidyverse.org/)) or interactively ([plotly](https://plotly.com/r/)) and can be customized with the corresponding functions of the respective visualization package.

## Installation

The latest released version of *weibulltools* from [CRAN](https://cran.r-project.org/web/packages/weibulltools/) can be installed with:

```{r, eval = FALSE}
install.packages("weibulltools")
```

### Development version
Install the development version of *weibulltools* from [GitHub](https://github.com/Tim-TU/weibulltools/) to use new features or to get a bug fix.

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("Tim-TU/weibulltools")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(weibulltools)
## basic example code
```

## Getting help

If you notice a bug or have suggestions for improvements, please submit an issue with a minimal reproducible example on [GitHub](https://github.com/Tim-TU/weibulltools/issues). For further questions, please contact [Tim-Gunnar Hensel](mailto:tim-gunnar.hensel@tu-berlin.de). 
