---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(tibble.print_min = 5, tibble.print_max = 5)
```

# weibulltools

<!-- badges: start -->
[![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![CRAN status](https://www.r-pkg.org/badges/version/weibulltools)](https://CRAN.R-project.org/package=weibulltools)
<!-- badges: end -->

## Overview

The *weibulltools* package focuses on statistical methods and visualizations that are often used in reliability engineering.  

The goal of *weibulltools* is to equip the user with a compact and easily accessible set of methods and visualization tools that make the examination and adjustment as well as the analysis and interpretation of field data (and bench tests) as simple as possible.  

Besides the well-known weibull analysis, *weibulltools* supports multiple lifetime distributions and also contains Monte Carlo methods for the correction and completion of imprecisely recorded or unknown lifetime characteristics.  

Plots are created statically ([ggplot2](https://ggplot2.tidyverse.org/)) or interactively ([plotly](https://plotly.com/r/)) and can be customized with the corresponding functions of the respective visualization package.

## Installation

The latest released version of *weibulltools* from [CRAN](https://cran.r-project.org/web/packages/weibulltools/) can be installed with:

```{r, eval = FALSE}
install.packages("weibulltools")
```

### Development version
Install the development version of *weibulltools* from [GitHub](https://github.com/Tim-TU/weibulltools/) to use new features or to get a bug fix.

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("Tim-TU/weibulltools")
```

## Usage

### Getting started
Create consistent reliability data with columns:  

* `x` - lifetime characteristic  
* `status` - binary data (0 for censored units and 1 for failed units)
* `id` (optional) - identifier for units  

```{r, message = FALSE}
library(weibulltools)

rel_tbl <- reliability_data(data = shock, x = distance, status = status)
rel_tbl
```

### Probability estimation and visualization
Estimation of failure probabilities using different non-parametric methods: 

```{r, probability estimation}
prob_tbl <- estimate_cdf(x = rel_tbl, methods = c("mr", "kaplan", "johnson", "nelson"))
prob_tbl
```

Visualization of the obtained results in a distribution-specific probability plot: 

```{r, probability visualization}
prob_vis <- plot_prob(x = prob_tbl, distribution = "weibull", 
                      title_main = "Weibull Probability Plot", 
                      title_x = "Distance", 
                      title_y = "Probability of Failure in %", 
                      title_trace = "Methods",
                      plot_method = "gg")
prob_vis
```

### Model estimation and visualization
Parametric model estimation with respect to the used methods: 

```{r, model estimation}
rr_list <- rank_regression(x = prob_tbl, distribution = "weibull")
rr_list
```

Model visualization in an existing probability plot:

```{r, model visualization}
mod_vis <- plot_mod(p_obj = prob_vis, x = rr_list, distribution = "weibull", 
                    title_trace = "RR") 
mod_vis
```

## Getting help

If you notice a bug or have suggestions for improvements, please submit an issue with a minimal reproducible example on [GitHub](https://github.com/Tim-TU/weibulltools/issues). For further questions, please contact [Tim-Gunnar Hensel](mailto:tim-gunnar.hensel@tu-berlin.de). 
