---
title: "Life Data Analysis Part III - Mixture Models"
subtitle: "Segmented Regression and EM Algorithm"
author: 
  - "Tim-Gunnar Hensel" 
  - "David Barkemeyer"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_height: 6
    fig_width: 7
    fig_caption: yes
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Life Data Analysis Part III - Mixture Models}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  screenshot.force = FALSE,
  comment = "#>"
)
library(weibulltools)
```

In this vignette two methods for the separation of mixture models are presented. 
A mixture model can be assumed, if the points in a probability plot show one or 
more changes in slope, depict one or several saddle points or follow an S-shape. 
A mixed distribution often represents the combination of multiple failure modes 
and thus must be split in its components to get reasonable results in further analyses. 

Segmented regression aims to detect breakpoints in the sample data from whom a 
split in subgroups can be made. The EM-Algorithm is a computation-intensive method 
that iteratively tries to maximize a likelihood function, which is weighted by 
posterior probabilities, conditional probabilities that an observation belongs to subgroup _k_.  

In the following, the focus is on the application of these methods and their 
visualizations using functions `mixmod_regression()`, `mixmod_em()`, `plot_prob()` 
and `plot_mod()`. 

## Data: Voltage Stress Test

To apply the introduced methods a dataset `voltage` is used. The dataset contains 
observations for units that were passed to a high voltage stress test. _hours_ 
indicates the number of hours until a failure occurs or the number of hours until 
a unit was taken out of the test and has not failed. _status_ is a flag variable 
and describes the condition of a unit. If a unit has failed the flag is 1 and 0 
otherwise. The dataset is taken from _Reliability Analysis by Failure Mode_ [^note1]. 

[^note1]: Doganaksoy, N.; Hahn, G.; Meeker, W. Q.: _Reliability Analysis by Failure Mode_, 
          Quality Progress, 35(6), 47-52, 2002  
          
For consistent handling of the data, `weibulltools` introduces the function 
`reliability_data()` that converts the original dataset into a `wt_reliability_data` 
object. This formatted object allows to easily apply the presented methods.  

```{r dataset_voltage, message = FALSE}
voltage_tbl <- reliability_data(data = voltage, x = hours, status = status)
voltage_tbl
```

## Probability Plot for Voltage Stress Test Data

To get an intuition whether one can assume the presence of a mixture model, a 
Weibull probability plot is constructed. 

```{r probability plot weibull, fig.cap = "Figure 1: Plotting positions in Weibull grid.", message = FALSE}
# Estimating failure probabilities: 
voltage_cdf <- estimate_cdf(voltage_tbl, "johnson")

# Probability plot: 
weibull_plot <- plot_prob(
  voltage_cdf,
  distribution = "weibull", 
  title_main = "Weibull Probability Plot", 
  title_x = "Time in Hours", 
  title_y = "Probability of Failure in %",
  title_trace = "Defectives",
  plot_method = "ggplot2"
)

weibull_plot
```

<br>
Since there is an obvious slope change in the Weibull probability plot of _Figure 1_, 
the appearance of a mixture model is strengthened.  

## Segmented Regression with Package `weibulltools`

The method of segmented regression is implemented in the function `mixmod_regression()`. 
If a breakpoint was detected, the failure data is separated by that point. After 
breakpoint detection the function `rank_regression()` is called inside `mixmod_regression()` 
and is used to estimate the distribution parameters of the subgroups. The visualization 
of the obtained results is done by functions `plot_prob()` and `plot_mod()`.  

```{r segmented weibull I, fig.cap = "Figure 2: Subgroup-specific plotting positions using segmented regression.", message = FALSE}
# Applying mixmod_regression(): 
mixreg_weib <- mixmod_regression(
  x = voltage_cdf, 
  distribution = "weibull", 
  control = segmented::seg.control(nonParam = FALSE)
)

# Using plot_prob_mix(). 
mix_reg_plot <- plot_prob(
  x = mixreg_weib, 
  title_main = "Weibull Mixture Regression", 
  title_x = "Time in Hours", 
  title_y = "Probability of Failure", 
  title_trace = "Subgroup",
  plot_method = "ggplot2"
)

mix_reg_plot
```

<br>

```{r segmented weibull II, fig.cap = "Figure 3: Subgroup-specific regression lines using segmented regression.", message = FALSE}
# Using plot_mod() to visualize regression lines of subgroups: 
mix_reg_lines <- plot_mod(
  mix_reg_plot, 
  x = mixreg_weib, 
  title_trace = "Fitted Line"
)

mix_reg_lines
```

<br>
Without specifying the number of mixed components _(k)_ this method has separated 
the data in three groups. This can bee seen in _Figure 2_ and _Figure 3_. 
To sum up, an upside of this function is that one does not have to specify the 
number of mixing components, since segmentation happens in an automated fashion. 
Nevertheless the intention of this function is to give a hint for the existence of 
a mixture model. An in-depth analysis should be done afterwards.  

## EM Algorithm with Package `weibulltools`

The EM algorithm can be applied through the usage of the function `mixmod_em()`. 
In comparison to `mixmod_regression()` one has to specify _k_, the number of subgroups. 
The obtained results can be also visualized by functions `plot_prob()` and `plot_mod()`.  

```{r em weibull I, fig.cap = "Figure 4: Subgroup-specific plotting positions using EM algorithm.", message = FALSE}
# Applying mixmod_regression(): 
mix_em_weib <- mixmod_em(
  x = voltage_tbl, 
  distribution = "weibull",
  k = 2
)

# Using plot_prob(): 
mix_em_plot <- plot_prob(
  x = mix_em_weib,
  title_main = "Weibull Mixture EM", 
  title_x = "Time in Hours", 
  title_y = "Probability of Failure", 
  title_trace = "Subgroup",
  plot_method = "ggplot2"
)

mix_em_plot
```

```{r em weibull II, fig.cap = "Figure 5: Subgroup-specific regression lines using EM algorithm.", message = FALSE}

# Using plot_mod() to visualize regression lines of subgroups: 
mix_em_lines <- plot_mod(
  mix_em_plot, 
  x = mix_em_weib, 
  title_trace = "Fitted Line"
)

mix_em_lines
```

<br>
In comparison to `mixmod_regression()` the EM-Algorithm can also assign censored 
items to a specific subgroup. Hence, an individual analysis of the mixing components, 
depicted in _Figure 4_ and _Figure 5_, is possible. In conclusion an analysis of 
a mixture model using `mixmod_em()` is statistically founded. A drawback of this 
function is, that the identification of the number of subgroups can not be determined automatically.
