---
title: "Life Data Analysis Part IV - Mixture Models"
subtitle: "Segmented Regression and EM Algorithm"
author: 
  - "Tim-Gunnar Hensel" 
  - "David Barkemeyer"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_height: 6
    fig_width: 7
    fig_caption: yes
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Life Data Analysis Part IV - Mixture Models}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  screenshot.force = FALSE,
  comment = "#>"
)
library(weibulltools)
```

In this vignette two methods for the separation of mixture models are presented. 
A mixture model can be assumed, if the points in a probability plot show one or 
more changes in slope, depict one or several saddle points or follow an S-shape. 
A mixed distribution often represents the combination of multiple failure modes 
and thus must be split in its components to get reasonable results in further analyses. 

Segmented regression aims to detect breakpoints in the sample data from which a 
split in subgroups can be made. The expectation-maximization (EM) algorithm is a 
computation-intensive method that iteratively tries to maximize a likelihood function, 
which is weighted by posterior probabilities. These are conditional probabilities 
that an observation belongs to subgroup _k_.  

In the following, the focus is on the application of these methods and their 
visualizations using the functions `mixmod_regression()`, `mixmod_em()`, 
`plot_prob()` and `plot_mod()`. 

## Data: Voltage Stress Test

To apply the introduced methods the dataset `voltage` is used. The dataset contains 
observations for units that were passed to a high voltage stress test. _hours_ 
indicates the number of hours until a failure occurs or the number of hours until 
a unit was taken out of the test and has not failed. _status_ is a flag variable 
and describes the condition of a unit. If a unit has failed the flag is 1 and 0 
otherwise. The dataset is taken from _Reliability Analysis by Failure Mode_ [^note1]. 

[^note1]: Doganaksoy, N.; Hahn, G.; Meeker, W. Q.: _Reliability Analysis by Failure Mode_, 
          Quality Progress, 35(6), 47-52, 2002  
          
For consistent handling of the data, {weibulltools} introduces the function 
`reliability_data()` that converts the original dataset into a `wt_reliability_data` 
object. This formatted object allows to easily apply the presented methods.  

```{r dataset_voltage, message = FALSE}
voltage_tbl <- reliability_data(data = voltage, x = hours, status = status)
voltage_tbl
```

## Probability Plot for Voltage Stress Test Data

To get an intuition whether one can assume the presence of a mixture model, a 
Weibull probability plot is constructed. 

```{r probability_plot_weibull, fig.cap = "Figure 1: Plotting positions in Weibull grid.", message = FALSE}
# Estimating failure probabilities: 
voltage_cdf <- estimate_cdf(voltage_tbl, "johnson")

# Probability plot: 
weibull_plot <- plot_prob(
  voltage_cdf,
  distribution = "weibull", 
  title_main = "Weibull Probability Plot", 
  title_x = "Time in Hours", 
  title_y = "Probability of Failure in %",
  title_trace = "Defectives",
  plot_method = "ggplot2"
)

weibull_plot
```

<br>
Since there is one obvious slope change in the Weibull probability plot of _Figure 1_, 
the appearance of a mixture model consisting of two subgroups is strengthened.  

## Segmented Regression with Package {weibulltools}

The method of segmented regression is implemented in the function `mixmod_regression()`. 
If a breakpoint was detected, the failure data is separated by that point. After 
breakpoint detection the function `rank_regression()` is called inside `mixmod_regression()` 
and is used to estimate the distribution parameters of the subgroups. The visualization 
of the obtained results is done by functions `plot_prob()` and `plot_mod()`.  

```{r segmented_weibull_I, fig.cap = "Figure 2: Subgroup-specific plotting positions using segmented regression.", message = FALSE}
# Applying mixmod_regression(): 
mixreg_weib <- mixmod_regression(
  x = voltage_cdf, 
  distribution = "weibull", 
  k = 2
)

mixreg_weib

# Using plot_prob_mix(). 
mix_reg_plot <- plot_prob(
  x = mixreg_weib, 
  title_main = "Weibull Mixture Regression", 
  title_x = "Time in Hours", 
  title_y = "Probability of Failure", 
  title_trace = "Subgroup",
  plot_method = "ggplot2"
)

mix_reg_plot
```

<br>

```{r segmented_weibull_II, fig.cap = "Figure 3: Subgroup-specific regression lines using segmented regression.", message = FALSE}
# Using plot_mod() to visualize regression lines of subgroups: 
mix_reg_lines <- plot_mod(
  mix_reg_plot, 
  x = mixreg_weib, 
  title_trace = "Fitted Line"
)

mix_reg_lines
```

<br>
The method has separated the data into $k = 2$ subgroups. This can bee seen in 
_Figure 2_ and _Figure 3_.  
An upside of this function is that the segmentation is done in a comprehensible 
manner.  

Furthermore, the segmentation process can be done automatically by setting `k = NULL`. 
The danger here, however, is an overestimation of the breakpoints.  

To sum up, this function should give an intention of the existence of a mixture 
model. An in-depth analysis should be done afterwards.  

## EM Algorithm with Package {weibulltools}

The EM algorithm can be applied through the usage of the function `mixmod_em()`. 
In contrast to `mixmod_regression()`, this method does not support an automatic 
separation routine and therefore _k_, the number of subgroups, must always be specified.  

The obtained results can be also visualized by the functions `plot_prob()` and `plot_mod()`.  

```{r em_weibull_I, fig.cap = "Figure 4: Subgroup-specific plotting positions using EM algorithm.", message = FALSE}
# Applying mixmod_regression(): 
mix_em_weib <- mixmod_em(
  x = voltage_tbl, 
  distribution = "weibull",
  k = 2
)

mix_em_weib

# Using plot_prob(): 
mix_em_plot <- plot_prob(
  x = mix_em_weib,
  title_main = "Weibull Mixture EM", 
  title_x = "Time in Hours", 
  title_y = "Probability of Failure", 
  title_trace = "Subgroup",
  plot_method = "ggplot2"
)

mix_em_plot
```

```{r em_weibull_II, fig.cap = "Figure 5: Subgroup-specific regression lines using EM algorithm.", message = FALSE}

# Using plot_mod() to visualize regression lines of subgroups: 
mix_em_lines <- plot_mod(
  mix_em_plot, 
  x = mix_em_weib, 
  title_trace = "Fitted Line"
)

mix_em_lines
```

<br>
One advantage over `mixmod_regression()` is, that the EM algorithm can also assign 
censored items to a specific subgroup. Hence, an individual analysis of the mixing 
components, depicted in _Figure 4_ and _Figure 5_, is possible.  
In conclusion an analysis of a mixture model using `mixmod_em()` is statistically founded.
