---
title: "Life Data Analysis Part III - Confidence Intervals"
subtitle: "Estimation and visualization"
author: 
  - "Tim-Gunnar Hensel" 
  - "David Barkemeyer"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_height: 6
    fig_width: 7
    fig_caption: yes
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Life Data Analysis Part III - Confidence Intervals}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  screenshot.force = FALSE,
  comment = "#>"
)
library(weibulltools)
library(plotly)
```

Computing confidence intervals is a well-known measure to express the uncertainty associated with an estimation. They can be 

## Fisher's confidence bounds

Maximum Likelihood estimators are asymptotically normally distributed. Therefore confidence intervals can be approximated using the normal distribution. Let $\vartheta$ be an unknown distribution parameter and $\hat{\vartheta}_{\text{MLE}}$ be an Maximum Likelihood estimator for $\vartheta$ with estimated standard error $\sqrt{\hat{Var}(\hat{\vartheta})}$. Then 
$$\bigg[\vartheta_{\text{lower}} = \hat{\vartheta}_{\text{MLE}} - z_{1 - \frac{\alpha}{2}} \sqrt{\hat{Var}(\hat{\vartheta}_{\text{MLE}})} \, ; \quad \vartheta_{\text{upper}} = \hat{\vartheta}_{\text{MLE}} + z_{1 - \frac{\alpha}{2}} \sqrt{\hat{Var}(\hat{\vartheta}_{\text{MLE}})} \bigg]$$
is the approximated two-sided confidence interval for $\vartheta$, where $\alpha$ is the confidence level and $z_p$ is the $p$-quantile of the standard normal distribution. The standard error is estimated using the so-called *delta method*.

## Data

To calculate the introduced confidence intervals the `shock` datasets is used. In this dataset kilometer-dependent problems that have occurred on shock absorbers are reported. In addition to failed items the dataset also contains non-defectives (*censored*) observations. The data can be found in _Statistical Methods for Reliability Data_ [^note4]. 

[^note4]: Meeker, W. Q.; Escobar, L. A.: _Statistical Methods for Reliability Data_, 
          _New York, Wiley series in probability and statistics_, 1998, p. 630  
          
For consistent handling of the data, {weibulltools} introduces the function `reliability_data()` 
that converts the original dataset into a `wt_reliability_data` object. This formatted object 
allows to easily apply the presented methods.  

```{r dataset_shock, message = FALSE}
shock_tbl <- reliability_data(data = shock, x = distance, status = status)
shock_tbl
```       

## Confidence intervals with {weibulltools}

Before calculating confidence intervals with {weibulltools} we need to conduct the basic steps of the Weibull analysis as described in the previous vignettes. 

```{r}
# Estimation of failure probabilities
shock_cdf <- estimate_cdf(shock_tbl, methods = "johnson")

# Rank Regression
shock_rr <- rank_regression(shock_cdf, distribution = "weibull")

# Maximum Likelihood Estimation 
shock_mle <- ml_estimation(shock_tbl, distribution = "weibull")
```

### Confidence intervals for distribution parameters

The confidence intervals for the estimated distribution parameters are included in the model output of `rank_regression()` and `ml_estimation()`, respectively.

```{r}
shock_rr$confint

shock_mle$confint
```

The `confint` element of the model output is a matrix with the parameter names as row names and the confidence bounds as column names. Different confidence levels can be specified using the argument `conf_level`.

```{r}
shock_rr_99 <- rank_regression(shock_cdf, distribution = "weibull", conf_level = 0.99)
shock_rr_99$confint
```

### Calculating confidence bounds for failure probabilities

Confidence bounds for failure probabilities are calculated using either `confint_betabinom()` or `confint_fisher()`. As explained in the introduction of this vignette the betabinom confidence bounds are only applicable to the output of `rank_regression()` whereas the fisher confidence bounds are only applicable to the output of `ml_estimation()`.

```{r}
conf_bb <- confint_betabinom(shock_rr)
conf_bb

conf_fisher <- confint_fisher(shock_mle)
conf_fisher
```

The outputs of both functions contain the calculated bounds for 100 equidistant quantiles ranging from the minimum to the maximum observed failure. In addition, confidence bounds for selected probabilities can be included by specifiying the argument `b_lives`.

### Visualizing confidence bounds for failure probabilities

Confidence bounds for failure probabilities can be added to a probability plot with `plot_conf()`.

```{r}
# Probability plot
weibull_grid <- plot_prob(shock_cdf, distribution = "weibull")

weibull_conf_bb <- plot_conf(weibull_grid, conf_bb)
weibull_conf_bb

weibull_conf_fisher <- plot_conf(weibull_grid, conf_fisher)
weibull_conf_fisher
```

As you can see, `plot_conf()` not only adds the confidence limits to the probability plot, but also includes the estimated linearized CDF. There is no need to call `plot_mod()`. In fact, the same routines used by `plot_mod()` are called under the hood. This ensures that no confidence bounds are drawn without a model and model line is based on the same points as the confidence bounds.
