---
title: "Life Data Analysis Part I - Estimation of Failure Probabilities"
subtitle: "A Non-parametric Approach"
author: 
  - "Tim-Gunnar Hensel" 
  - "David Barkemeyer"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_height: 6
    fig_width: 7
    fig_caption: yes
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Life Data Analysis Part I - Estimation of Failure Probabilities}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  screenshot.force = FALSE,
  comment = "#>"
)
library(weibulltools)
```

This document presents non-parametric estimation methods for the computation of 
failure probabilities of complete data (failures) taking (multiple) right-censored 
units into account. A unit can be either a single component, an assembly or an entire system.  

Further, the estimation results are presented in distribution-specific probability plots.

## Introduction to Life Data Analysis

If the lifetime (or any other damage-equivalent quantity such as distance or 
load cycles) of a unit is considered to be a continuous random variable _T_, 
then the probability that a unit has failed at _t_ is defined by its 
_CDF (cumulative distribution function)_ _F(t)_.

$$ P(T\leq t) = F(t) $$

In order to obtain an estimate of the _CDF_ for each observation $t_1, t_2, ..., t_n$ 
two approaches are possible. Using a parametric lifetime distribution requires that 
the underlying assumptions for the sample data are valid. If the distribution-specific 
assumptions are correct, the model parameters can be estimated and the _CDF_ is 
computable. But if conditions could not be met, interpretations and derived conclusions 
are not reliable.

A more general approach for the calculation of cumulative failure probabilities is 
to use non-parametric statistical estimators $\hat{F}(t_1), \hat{F}(t_2), ..., \hat{F}(t_n)$. 
In comparison to a parametric distribution no general assumptions must be held. 
For non-parametric estimators, an ordered sample of size $n$ is needed. Starting 
at $1$, the ranks $i \in \{1, 2, ..., n \}$ are assigned to the ascending sorted 
sample values. Since there is a known relationship between ranks and corresponding 
ranking probabilities a _CDF_ can be determined.  

But rank distributions are systematically skewed distributions and thus the median 
value instead of the expected value $E\left[F\left(t_i\right)\right] = \frac{i}{n + 1}$ 
is used for the estimation [^note1]. This skewness is visualized in _Figure 1_. 

[^note1]: Kapur, K. C.; Lamberson, L. R.: _Reliability in Engineering Design_, 
          _New York: Wiley_, 1977, pp. 297-301  

```{r rank densities, fig.cap = "Figure 1: Densities for different ranks i in samples of size n = 10.", message = FALSE}
library(dplyr) # data manipulation 
library(ggplot2) # visualization

x <- seq(0, 1, length.out = 100) # CDF
n <- 10 # sample size
i <- c(1, 3, 5, 7, 9) # ranks
r <- n - i + 1 # inverse ranking

df_dens <- expand.grid(cdf = x, i = i) %>% 
  mutate(n = n, r = n - i + 1, pdf = dbeta(x = x, shape1 = i, shape2 = r))

densplot <- ggplot(data = df_dens, aes(x = cdf, y = pdf, colour = as.factor(i))) + 
  geom_line() + 
  scale_colour_discrete(guide = guide_legend(title = "i")) + 
  theme_bw() + 
  labs(x = "Failure Probability", y = "Density")
densplot
```

### Failure Probability Estimation  

In practice, a simplification for the calculation of the median value, also called 
median rank, is made. The formula of _Benard's_ approximation is given by 

$$\hat{F}(t_i) \approx \frac{i - 0,3}{n + 0,4} $$ 

and is described in _The Plotting of Observations on Probability Paper_ [^note2]. 

[^note2]: Benard, A.; Bos-Levenbach, E. C.: _The Plotting of Observations on Probability Paper_, 
          _Statistica Neerlandica 7 (3)_, 1953, pp. 163-173  
          
However, this equation only provides valid estimates for failure probabilities if 
all units in the sample are defectives (`estimate_cdf(methods = "mr", ...)`).

In field data analysis, however, the sample mainly consists of intact units and 
only a small fraction of units failed. Units that have no damage at the point of 
analysis and also have not reached the operating time or mileage of units that 
have already failed, are potential candidates for future failures.  As these, 
for example, still are likely to fail during a specific time span, like the 
guarantee period, the _CDF_ must be adjusted upwards by these potential candidates.  

A commonly used method for correcting probabilities of (multiple) right-censored 
data is _Johnson's_ method (`estimate_cdf(methods = "johnson", ...)`). By this method, 
all units that are included in the period looked at are sorted in an ascending order 
of their operating time (or any other damage-equivalent quantity). If there are units 
that have not failed before the _i_-th failure, an adjusted rank for the _i_-th failure 
is formed. This correction takes the potential candidates into account and increases 
the rank number. In consequence, a higher rank leads to a higher failure probability. 
This can be seen in _Figure 1_.
  
The rank adjustment is determined with: 

$$j_i = j_{i-1} + x_i \cdot I_i, \;\; with \;\; j_0 = 0$$

Here, $j_ {i-1}$ is the adjusted rank of the previous failure, $x_i$ is the number 
of defectives at $t_i$ and $I_i$ is the increment that corrects the considered rank 
by the potential candidates. 

$$I_i=\frac{(n+1)-j_{i-1}}{1+(n-n_i)}$$

The sample size is $n$ and $n_i$ is the number of units that have a lower $t$ than 
the _i_-th unit. Once the adjusted ranks are calculated, the _CDF_ can be estimated 
according to _Benard's_ approximation.  

Other methods in `weibulltools` that can also handle (multiple) right-censored data 
are the _Kaplan-Meier_ estimator (`estimate_cdf(methods = "kaplan")`) and the 
_Nelson-Aalen_ estimator (`estimate_cdf(methods = "nelson")`). 

### Probability Plotting  

After computing failure probabilities a method called _probability plotting_ is 
applicable. It is a graphical _goodness of fit_ technique that is used in assessing 
whether an assumed distribution is appropriate to model the sample data.  

The axes of a probability plot are transformed in such a way that the _CDF_ of a 
specified model is represented through a straight line. If the plotted points 
(`plot_prob()`) lie on an approximately straight line it can be said that the 
chosen distribution is adequate.  

The two-parameter Weibull distribution can be parameterized with parameters $\mu$ 
and $\sigma$ such that the _CDF_ is characterized by the following equation:  

$$F(t)=\Phi_{SEV}\left(\frac{\log(t) - \mu}{\sigma}\right)$$

The advantage of this representation is that the Weibull is part of the 
(log-)location-scale family. A linearized representation of this _CDF_ is: 

$$\Phi^{-1}_{SEV}\left[F(t)\right]=\frac{1}{\sigma} \cdot \log(t) - \frac{\mu}{\sigma}$$

This leads to the following transformations regarding the axes: 

* Abscissa: $x = \log(t)$ 
* Ordinate: $y = \Phi^{-1}_{SEV}\left[F(t)\right]$, which is the quantile function 
  of the SEV (_smallest extreme value_) distribution and can be written out with 
  $\log\left\{-\log\left[1-F(t)\right]\right\}$.  

Another version of the Weibull _CDF_ with parameters $\eta$ and $\beta$ results in 
a _CDF_ that is defined by the following equation:  

$$F(t)=1-\exp\left[ -\left(\frac{t}{\eta}\right)^{\beta}\right]$$

Then a linearized version of the CDF is: 

$$ \log\left\{-\log\left[1-F(t)\right]\right\} = \beta \cdot \log(t) - \beta \cdot \log(\eta)$$

Transformations regarding the axes are: 

* Abscissa: $x = \log(t)$ 
* Ordinate: $y = \log\left\{-\log\left[1-F(t)\right]\right\}$.
  
It can be easily seen that the parameters can be converted into each other. The 
corresponding equations are: 

$$\beta = \frac{1}{\sigma}$$  

and 

$$\eta = \exp\left(\mu\right).$$   

## Data: Shock Absorber

To apply the introduced methods of non-parametric failure probability estimation 
and probability plotting the `shock` data is used. In this dataset kilometer-dependent 
problems that have occurred on shock absorbers are reported. In addition to failed 
items the dataset also contains non-defectives (*censored*) observations. The data 
can be found in _Statistical Methods for Reliability Data_ [^note3]. 

[^note3]: Meeker, W. Q.; Escobar, L. A.: _Statistical Methods for Reliability Data_, 
          _New York, Wiley series in probability and statistics_, 1998, p. 630
          
For consistent handling of the data, `weibulltools` introduces the function 
`reliability_data()` that converts the original dataset into a `wt_reliability_data` 
object. This formatted object allows to easily apply the presented methods.  

```{r dataset shock, message = FALSE}
shock_tbl <- reliability_data(data = shock, x = distance, status = status)
shock_tbl
```

## Estimation of Failure Probabilities with Package `weibulltools`

First, we are interested in how censored observations influence the estimation of 
failure probabilities in comparison to the case where only failed units are considered. 
To deal with survived and failed units we will use `estimate_cdf()` with 
`methods = "johnson"`, whereas `methods = "mr"` only considers failures. 

```{r failure probabilities}
# Estimate CDF with both methods: 
cdf_tbl <- estimate_cdf(shock_tbl, methods = c("mr", "johnson"))

# First case where only failed units are taken into account:
cdf_tbl_mr <- cdf_tbl %>% filter(method == "mr")
cdf_tbl_mr

# Second case where both, survived and failed units are considered:
cdf_tbl_john <- cdf_tbl %>% filter(method == "johnson") 
cdf_tbl_john
```

<br>

If we compare both outputs we can see that survivors reduce the probabilities. 
But this is just that what was expected since undamaged units with longer or equal 
lifetime characteristic _x_ let us gain confidence in the product. 

## Probability Plotting with Package `weibulltools`

The estimated probabilities should now be presented in a probability plot. 
With `plot_prob()` probability plots for several lifetime distributions can be constructed 
and estimates of multiple methods can be displayed at once. 

### Weibull Probability Plot

```{r probability plot weibull, fig.cap = "Figure 3: Plotting positions in Weibull grid.", message = FALSE}
# Weibull grid for estimated probabilities: 
weibull_grid <- plot_prob(
  cdf_tbl,
  distribution = "weibull", 
  title_main = "Weibull Probability Plot", 
  title_x = "Mileage in km", 
  title_y = "Probability of Failure in %",
  title_trace = "Method",
  plot_method = "ggplot2"
)

weibull_grid
```

<br>

_Figure 3_ shows that the consideration of survivors (orange points, _Method: johnson_) 
decreases the failure probability in comparison to the sole evaluation of failed 
items (green points, _Method: mr_).  

### Log-normal Probability Plot

Finally, we want to use a log-normal probability plot to visualize the estimated 
failure probabilities.

```{r probability plot log-normal, fig.cap = "Figure 4: Plotting positions in log-normal grid.", message = FALSE}
# Log-normal grid for estimated probabilities: 
lognorm_grid <- plot_prob(
  cdf_tbl,
  distribution = "lognormal",
  title_main = "Log-normal Probability Plot",
  title_x = "Mileage in km",
  title_y = "Probability of Failure in %",
  title_trace = "Method",
  plot_method = "ggplot2"
)

lognorm_grid
```

<br>

On the basis of _Figure 3_ and _Figure 4_ we can subjectively assess the goodness 
of fit of Weibull and log-normal. It can be seen that in both grids, the plotted 
points roughly fall on a straight line. Hence one can say that the Weibull as well 
as the log-normal are good model candidates for the `shock` data.
